---
title: "fitQuasar2CR: Cox-Reid + EB Beta-Binomial Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fitQuasar2CR: Cox-Reid + EB Beta-Binomial Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4
)
```

# Overview

`fitQuasar2CR` is a beta-binomial regression for allele-specific count data
that:

1. parameterizes the BB precision as $\psi = \log(1/M)$,
2. estimates $\psi$ per SNP with the **Cox-Reid adjusted profile likelihood**,
3. **moderates** $\psi$ across SNPs with a smooth trend in `log(baseMean)` +
   empirical-Bayes shrinkage on a Normal prior, and
4. exposes a `model.matrix` interface so `testCoef`, `testContrast`, and
   `testLRT` work like in `limma` / `edgeR` / `DESeq2`.

The full math, derivations of the score and information matrix, and the
empirical-Bayes prior derivation are in `inst/doc/quasar2CR_math.md` (also
shipped at the top of the QuASAR2ex/ folder in the repo).

# A quick example

```{r example, eval = FALSE}
library(QuASAR2)
set.seed(1)

## simulate 50 SNPs with 8 reps each, half-treatment effect on trt=B
G   <- 50
n_g <- 8
trt <- factor(rep(c("A","A","A","A","B","B","B","B"), length.out = n_g))
M_g <- exp(runif(G, log(20), log(200)))

dat <- do.call(rbind, lapply(seq_len(G), function(g) {
  ## true effect: half the SNPs have b_trt = 1, half are null
  b_trt <- if (g <= G/2) 1.0 else 0
  mu  <- 1 / (1 + exp(-(b_trt * (trt == "B"))))
  pp  <- rbeta(n_g, M_g[g] * mu, M_g[g] * (1 - mu))
  N   <- sample(40:150, n_g, replace = TRUE)
  R   <- rbinom(n_g, N, pp); A <- N - R
  data.frame(identifier = sprintf("snp_%03d", g), R = R, A = A, trt = trt)
}))

fit <- fitQuasar2CR(dat, design = ~ trt, verbose = FALSE)
print(fit)
```

# Inference

## Wald test for a single coefficient

```{r wald, eval = FALSE}
res_wald <- testCoef(fit, coef = "trtB")
head(res_wald)
```

By default the reference distribution is Student-t with `df = n_g - p_g`
("residual" df).  For closer-to-nominal Type-I error at small `n_g`, pass
`df_method = "moderated"` which uses `df = n_g - p_g + 1/sigma2_psi` (the EB
prior precision); see the calibration study in `QuASAR2ex/` for the numbers.

```{r moderated, eval = FALSE}
res_wald_mod <- testCoef(fit, coef = "trtB", df_method = "moderated")
```

## Likelihood ratio test against a reduced design

```{r lrt, eval = FALSE}
res_lrt <- testLRT(fit, reduced = ~ 1)
head(res_lrt)
```

This computes $\Lambda_g = 2(\ell_{\rm full} - \ell_{\rm red})$ holding
dispersion fixed at the full-model MAP estimate (edgeR convention).

## Arbitrary linear contrasts

```{r contrasts, eval = FALSE}
L <- makeContrastsQuasar("trtB", fit)            # single-row contrast
res_F <- testContrast(fit, L)
```

For multi-treatment designs you can stack contrasts as rows of `L`.

# What's saved on the fit

```{r inspect, eval = FALSE}
fit$sigma2_psi              # EB prior variance on log(1/M)
head(fit$dispersion)        # per-SNP psi, M, psi_gw, psi_trend, var_samp, baseMean
fit$eps                     # MLE of shared sequencing error
fit$loglik                  # vector of per-SNP log-likelihoods at convergence
```

# When to use what

| Situation                            | `ref`      | `df_method`  |
|--------------------------------------|------------|--------------|
| Default, safe (publication FDR)      | `"t"`/`"F"`| `"residual"` |
| Maximum power, ~nominal Type-I       | `"t"`/`"F"`| `"moderated"`|
| Asymptotic comparison, large `n_g`   | `"normal"`/`"chisq"` | n/a |

The `"residual"` default is mildly conservative (Type-I ~ 0.035 across a
27-cell coverage x dispersion x n_g grid).  The `"moderated"` option gives
Type-I ~ 0.052 (nominal) but can mildly exceed 0.05 at very small `n_g`
combined with very weak overdispersion.
